Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array

### 2023-11-05 docs/manuscript/paper.tex
all comments addressed

For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. \newline

@@

\section{Kalman filtering} \label{sec:kalman}
In this appendix, we describe the Kalman filter algorithm used in this paper. The general recursion relations for the discrete-time Kalman filter applied to an arbitrary linear dynamical system. are outlined in Section \ref{sec_kalman_general}. The mapping onto our specific continuous-time state-space model described in Section \ref{sec:model} is given in Section \ref{sec_kalman_specific}.

\subsection{Recursion equations}\label{sec_kalman_general}
The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ the control matrix. $\boldsymbol{u}_k$ the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline

The Kalman filter is a recursive estimator with two distinct stages; a ``predict" stage and an ``update" stage. The predict stage predicts the estimate of the state at discrete step $k$, given the state estimates from step $k-1$. Specifically, the predict step proceeds as,
\begin{align}
	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ .
\end{align}
Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
\begin{align}
	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
\end{align}
Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.  $\boldsymbol{K}_k = \text{argmin} \left[ \boldsymbol{E}(\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k) \right ]$. For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}. \newline

To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$,$\boldsymbol{Y}_k$,$\boldsymbol{F}_k$,$\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$,$\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$. In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.

\subsection{Full state space representation}\label{sec_kalman_specific}
We apply the Kalman machinery of Section \ref{sec_kalman_general} onto the state-space model described in Section \ref{sec:model} as follows. \newline

We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e 
\begin{equation}
	\boldsymbol{X}(t) = \left(f_{\rm p}^{(1)}(t), f_{\rm p}^{(2)}(t), ..., f_{\rm p}^{(N)}(t)\right) \ .
\end{equation}
Analogously,  
\begin{equation}
	\boldsymbol{Y}(t) = \left(f_{\rm m}^{(1)}(t), f_{\rm m}^{(2)}(t), ..., f_{\rm m}^{(N)}(t) \right) \ .
\end{equation}
The states evolve according the continuous dynamical equation (c.f. Equation \eqref{eq:frequency_evolution})
\begin{equation}
	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \label{eq:kalmn2}
\end{equation}
where $\boldsymbol{A}$ is a diagonal $N \times N$ square matrix,
\begin{equation}
	\boldsymbol{A} = \text{diag} \left(\gamma^{(1)}, \gamma^{(2)}, ..., \gamma^{(N)}\right) \ ,
\end{equation}
and $\boldsymbol{C}(t)$ a time dependent vector with $n$-th component
\begin{equation}
	C^{(n)} =\gamma^{(n)} \left(f_{\rm em}^{(n)} (t_1) + \dot{f}_{\rm em}(t_1)^{(n)} t \right) +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
\end{equation}
The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, where
\begin{equation}
	\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right) \ .
\end{equation}

The evolution of the state is an Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, which has a general solution given by \citep{gardiner2009stochastic},
\begin{equation}
	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ . \label{eq:gardenier}
\end{equation} 
From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2} where
\begin{align}
	\boldsymbol{F}_k &= e^{\boldsymbol{A}\left(\Delta t \right)} \  \\
	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
\end{align}
\begin{align}
	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_i}^{t_{i+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ' ,
\end{align}
\begin{equation}
	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \label{eq:appendix_noise}
\end{equation}
where $ \Delta t = t_{k+1} - t_k$ and 
\begin{align}
	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left[\Delta t + t_k \right] \nonumber \\ 
	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ .
\end{align}
From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
\begin{equation}
	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
\end{equation}
for 
\begin{equation}
Q^{(n)} = \frac{- [\sigma^{n}]^2}{2 \gamma^{(n)}} \left( e^{-2 \gamma^{(n)} \Delta t} -1\right) \ .
\end{equation}

The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$. These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Specifically, 
$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}. The measurement covariance $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$.

\section{Fiducial pulsar values}\label{appendix_fiducial}
\textcolor{red}{TK: AM suggests a table of the pulsars used with there fiducial values. Do we really need this? The argument is for reproducibility, but we have a nice github repo that holds all the data we actually used, jupyter notebooks that show how the calculations are done. Perhaps it is sufficient to point the reader to the repo... MNRAS encourages a "data availability" section before Acknowledgements that points readers to code/data/github repository}

\section{Wasserstein distance}\label{sec:wasserstein}
In this section we describe the Wasserstein distance used in Section \ref{sec:multiple_noise}. In Section \ref{sec:appendix_overview_WD} we give a brief overview of the Wasserstein distance generally. In Section \ref{sec:appendix_WD_results} we present for reproducibility the Wasserstein distance calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}. This corresponds to the mean values reported in Table \ref{tab:Wasserstein}.

\subsection{Overview of the Wasserstein distance}\label{sec:appendix_overview_WD}

The Wasserstein distance is a metric that defines a distance between two probability distributions $\mu(x)$ and $\nu(x)$. It has an intuitive interpretation as the lowest total cost with which one can move probability mass from $\mu$ to $\nu$, with respect to a cost function $c(x,y)$. For this reason it is sometimes known as the "Earth mover's distance". The $p$-th order Wasserstein distance between two distributions is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \lambda (x,y)\right)^{1/p} \label{eq:wasserstein} \ ,
\end{eqnarray}
where $p \in [1,\infty)$, $\lambda(x,y)$ is the transport plan, and $\Gamma(\mu, \nu)$ is the set of all joint probability distributions for $(x,y)$ that have marginals $\mu$ and $\nu$, i.e. $\Gamma(\mu, \nu)$ is the set of couplings of $\mu$ and $\nu$.

The cost function can be freely chosen for the nature of the problem, but is commonly taken to be the absolute value function, $c(x,y) = |x-y|$, and we adopt the same cost function in this work. Given this choice of $c(x,y)$, for an empirical distribution $\hat{\mu}_n$ composed of $n$ i.i.d samples drawn from the underlying probability distribution $\mu$, the solution to the discrete form of Equation \eqref{eq:wasserstein} can be computed generally by the Hungarian algorithm \citep{Kuhn} in polynomial time $\mathcal{O}(n^3)$ \citep{Villani2009}. However, for $\mu$ defined on $\mathbb{R}^d$ it is well-known \citep{Dudley} that the Wasserstein distance suffers from the curse of dimensionality for dimensions $d \geq 3$:
\begin{eqnarray}
	\boldsymbol{E} [W_1(\hat{\mu}_n,\mu)] \asymp n^{-\frac{1}{d}} \ .
\end{eqnarray}
In order to circumvent this issue, in this work we calculate the Wasserstein distance exclusively between the one-dimensional marginalized posteriors, i.e. setting $d=1$. In this case Equation \eqref{eq:wasserstein} has a convenient, easily computed, closed form:
\begin{eqnarray}
	W_p(\mu,\nu)= \left(\int_0^1  F_{\mu}^{-1} (z) - F_{\nu}^{-1} (z) dz \right)^{1/p} \, \label{eq:wasserstein}
\end{eqnarray}
where $F_{\mu}(z)$ is the cumulative distribution function of $\mu$. \newline

The use of the Wasserstein distance over more common probability measures such as the Kullback–Leibler divergence or the  Kolmogorov-Smirnoff test has a number of advantages. Firstly, it is intuitive, being the minimal amount of work required to transform one distribution into another. It also inherits the units of the underlying distributions, providing a convenient heuristic; if $W_1 =1$ between a pair of distributions with units e.g. radians then we can interpret this as the cost to transform between the two distributions is 1 radian. Further interpretability is provided by the Monge-Kantorovich duality \citep{villani2003topics,Villani2009}:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \ , \label{eq:WDdefn}
\end{eqnarray}
where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value. That is, if $W_1 =1$ then the difference in the expected value of two random variables drawn from the two distributions is less than 1 radian. Secondly it is a versatile, non-parametric measure that requires no specific properties of the probability distributions that are being compared such as e.g. Gaussianity. Any two distributions can be compared, irregardless of whether they are continuous, discrete, singular or defined on different metric spaces. Thirdly, it is a true distance metric and as such satisfies desirable properties of a measure of distance such as symmetry and the triangle inequality. It also respects the geometry of the underlying metric space over which the distributions are defined. \newline

\subsection{Wasserstein distance for static parameters} \label{sec:appendix_WD_results}

Table \ref{tab:Wasserstein} summarizes the WD between the $5\times 10^5$ pairs of one-dimensional posteriors across the $10^3$ realizations, for each of the seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. In Figure \ref{fig:pairwise_wasserstein} we present the WD values for a subset ($=10^2$) of those noise realisations. We plot a subset of realisations so as not to overcrowd the figure and enable certain features in the figure to be resolved. \newline

The figure is composed of seven subplots, one for each element or ${\boldsymbol{\theta}}_{\rm gw}$. Each individual subplot is a lower triangular heat-map, where each point denotes the $W_1$ value between a pair of one-dimensional posteriors for that element. Smaller values of $W_1$ are magenta moving to through red to yellow for larger values of $W_1$. Note that the heat-map colour scaling is not the same for all subplots of the figure. \newline

>>>>

For each pulsar we adopt fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$, with the latter two quantities evaluated at the Solar System barycenter. A table of fiducial values is presented in Appendix \ref{appendix_fiducial} for the sake of reproducibility. The sky positions and colour-coded distances of the pulsars are displayed in Figure \ref{fig:pulsar_distrib}. The pulsar parameters are acquired via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. \newline

@@

\section{Kalman filter} \label{sec:kalman}
In this appendix, we describe the Kalman filter algorithm used in this paper. General recursion relations for the discrete-time Kalman filter are written down for an arbitrary linear dynamical system in Section \ref{sec_kalman_general}. The mapping onto the specific continuous-time state-space model in Section \ref{sec:model} is written down in Section \ref{sec_kalman_specific}.

\subsection{Recursion equations}\label{sec_kalman_general}
The linear Kalman filter operates on temporally discrete, noisy measurements $\boldsymbol{Y}_k$, which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$, via a linear transformation
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \ ,\label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise, $\mathcal{N} \sim (0,\boldsymbol{R}_k)$ with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter evolves the underlying states according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \ , \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}_k$ is the system dynamics matrix, $\boldsymbol{G}_k$ is the control matrix. $\boldsymbol{u}_k$ is the control vector, and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise, $\mathcal{N} \sim (0,\boldsymbol{Q}_k)$ with covariance $\boldsymbol{Q}_k$ \newline

The Kalman filter is a recursive estimator with two distinct stages: a ``predict" stage and an ``update" stage. The predict stage predicts $\hat{\boldsymbol{X}}_{k|k-1}$, the estimate of the state at discrete step $k$, given the state estimates from step $k-1$. Specifically, the predict step proceeds as
\begin{align}
	\hat{\boldsymbol{X}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{X}}_{k-1|k-1} + \boldsymbol{G}_k \boldsymbol{u}_k \ , \\
		\hat{\boldsymbol{P}}_{k|k-1} &=  \boldsymbol{F}_k \hat{\boldsymbol{P}}_{k-1|k-1} \boldsymbol{F}_k^\intercal + \boldsymbol{Q}_k  \ ,
\end{align}
where $\hat{\boldsymbol{P}}_{k|k-1}$ is the covariance of the prediction. Note that the predict stage is independent of the measurements. The measurement $\boldsymbol{Y}_k$ is included to update the prediction during the update stage as follows:
\begin{align}
	\boldsymbol{\epsilon}_{k} &= \boldsymbol{Y}_k - \boldsymbol{H}_k \hat{\boldsymbol{X}}_{k|k-1} \ , \\
	\boldsymbol{S}_k &= \boldsymbol{H}_k \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal + \boldsymbol{R}_k \ , \\
	\boldsymbol{K}_k &= \hat{\boldsymbol{P}}_{k|k-1} \boldsymbol{H}_k^\intercal \boldsymbol{S}_k^{-1} \ ,\label{eq:kalman gain} \\
		\hat{\boldsymbol{X}}_{k|k} &=\hat{\boldsymbol{X}}_{k|k-1} +\boldsymbol{K}_k  \boldsymbol{\epsilon}_{k}  \ , \label{eq:kalmangainupdate} \\
			\hat{\boldsymbol{P}}_{k|k} &= \left( \boldsymbol{I} - \boldsymbol{K}_k \boldsymbol{H}_k \right) 	\hat{\boldsymbol{P}}_{k|k-1} \ .
\end{align}
Equation \eqref{eq:kalman gain} defines the ``Kalman gain" $\boldsymbol{K}_k$ which is defined so as to minimise the mean squared error in the state estimate, i.e.  $\boldsymbol{K}_k = \text{argmin} \left \{ \boldsymbol{E}[ (\boldsymbol{X}_k - \hat{\boldsymbol{X}}_k)^2 ] \right \}$. For a full review of the Kalman filter, including its derivation, we refer the reader to \cite{Gelb:1974} and \cite{zarchan2000fundamentals}. \newline

To apply the Kalman filter in practice means specifying the eight component matrices that make up the ``machinery'' of the filter: $\boldsymbol{X}_k$, $\boldsymbol{Y}_k$, $\boldsymbol{F}_k$, $\boldsymbol{G}_k$, $\boldsymbol{u}_k$, $\boldsymbol{H}_k$, $\boldsymbol{Q}_k$ and $\boldsymbol{R}_k$. In Section \ref{sec_kalman_specific} we describe how the machinery is determined for the state-space model in Section \ref{sec:model}.

\subsection{State space representation of a PTA analysis}\label{sec_kalman_specific}
We apply the Kalman recursion relations in Section \ref{sec_kalman_general} to the state-space model of a PTA with $N$ pulsars described in Section \ref{sec:model} as follows. \newline

We identify $\boldsymbol{X}(t)$ with a vector of length $N$ composed of the intrinsic pulsar frequency states, i.e. 
\begin{equation}
	\boldsymbol{X}(t) = \left(f_{\rm p}^{(1)}(t), f_{\rm p}^{(2)}(t), ..., f_{\rm p}^{(N)}(t)\right) \ .
\end{equation}
Analogously,  we package the measured pulsar frequencies as
\begin{equation}
	\boldsymbol{Y}(t) = \left(f_{\rm m}^{(1)}(t), f_{\rm m}^{(2)}(t), ..., f_{\rm m}^{(N)}(t) \right) \ .
\end{equation}
The states evolve according to the continuous stochastic differential equation (c.f. Equation \eqref{eq:frequency_evolution})
\begin{equation}
	d \boldsymbol{X} = \boldsymbol{A} \boldsymbol{X} dt + \boldsymbol{C}(t) dt + \boldsymbol{\Sigma} d \boldsymbol{B}(t) \ , \label{eq:kalmn2}
\end{equation}
where $\boldsymbol{A}$ is a diagonal $N \times N$ matrix,
\begin{equation}
	\boldsymbol{A} = \text{diag} \left(-\gamma^{(1)}, -\gamma^{(2)}, ..., -\gamma^{(N)}\right) \ ,
\end{equation}
and $\boldsymbol{C}(t)$ is a time-dependent vector with $n$-th component
\begin{equation}
	C^{(n)} =\gamma^{(n)} \left[ f^{(n)} _{\rm em} (t_1) + \dot{f}^{(n)} _{\rm em}(t_1) \, t \right] +  \dot{f}_{\rm em}(t_1)^{(n)} \ .
\end{equation}
The $N \times N$ square matrix $\boldsymbol{\Sigma}$  governs the magnitude of the increments of Brownian motion (Wiener process) $d\boldsymbol{B}(t)$, with
\begin{equation}
	\boldsymbol{\Sigma} = \text{diag} \left(\sigma^{(1)}, \sigma^{(2)}, ..., \sigma^{(N)}\right) \ .
\end{equation}

In the idealized model above, each pulsar's rotational state evolves phenomenologically according to a mean-reverting Ornstein-Uhlenbeck process, described by a Langevin equation, Equation \eqref{eq:kalmn2}, whose general solution is given by \citep{gardiner2009stochastic},
\begin{equation}
	\boldsymbol{X}(t) = e^{\boldsymbol{A} t} \boldsymbol{X}(0) + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{C}(t') dt' + \int_0^t e^{\boldsymbol{A}(t-t')} \boldsymbol{\Sigma} d\boldsymbol{B}(t') \ . \label{eq:gardenier}
\end{equation} 
From Equation \eqref{eq:gardenier} we construct the discrete, recursive solution for $\boldsymbol{X}(t_k) = \boldsymbol{X}_k$ in the form of Equation \eqref{eq:kalman2}, with
\begin{align}
	\boldsymbol{F}_k &= e^{\boldsymbol{A} \Delta t } \  \\
	&= \text{diag}\left(e^{- \gamma^{(1)} \Delta t},e^{- \gamma^{(2)} \Delta t},...,e^{- \gamma^{(N)} \Delta t} \right) \ ,
\end{align}
\begin{align}
	\boldsymbol{G}_k \boldsymbol{u}_k &= \int_{t_k}^{t_{k+1}}  e^{\boldsymbol{A}\left( t_{k+1} - t' \right)}  \boldsymbol{C}(t') dt' \ , \\
	&= \left(G^{(1)}_k, G^{(2)}_k,...,G^{(N)}_k \right) ,
\end{align}
\begin{equation}
	\boldsymbol{w}_k = \int_{t_k}^{t_{k+1}} e^{\boldsymbol{A}\left( t_{k+1} - t' \right)} \boldsymbol{\Sigma} d \boldsymbol{B}(t') \ ,  \label{eq:appendix_noise}
\end{equation}
\begin{align}
	G_k^{(n)} =&    f^{(n)}_{\rm em}(t_1) + \dot{f}^{(n)}_{\rm em}(t_1)  \left(\Delta t + t_k \right) \nonumber \\ 
	&- e^{-\gamma \Delta t} \left[  f^{(n)}_{\rm em}(t_1) +\dot{f}^{(n)}_{\rm em}(t_1)  t_k \right] \ ,
\end{align}
and $\Delta t = t_{k+1} - t_k$. From Equation \eqref{eq:appendix_noise} the process noise covariance matrix is
\begin{equation}
	\boldsymbol{Q}_k \boldsymbol{\delta}_{kj}= \langle \boldsymbol{\eta}_k \boldsymbol{\eta}_j^\intercal \rangle = \text{diag} \left(Q^{(1)}, Q^{(2)},...,Q^{(N)}\right) \ ,
\end{equation}
with 
\begin{equation}
Q^{(n)} = \frac{[\sigma^{n}]^2}{2 \gamma^{(n)}} \left[ 1 - e^{-2 \gamma^{(n)} \Delta t}\right] \ .
\end{equation}

The two remaining unspecified component matrices of the Kalman filter are the measurement matrix $\boldsymbol{H}_k$ and the measurement covariance matrix $\boldsymbol{R}_k$. These are defined straightforwardly from Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Specifically, 
$\boldsymbol{H}_k$ is a diagonal matrix where the $n$-th component of the diagonal is given by $g^{(n)}(t_k)$ from Equation \eqref{eq:g_func}. The measurement covariance satisfies $\boldsymbol{R}_k = E \left[ \boldsymbol{v} \boldsymbol{v}^\intercal \right] = \sigma^2_{\rm m}$ for all $k$.

\section{Synthetic PTA specifications}\label{appendix_fiducial}
The synthetic PTA deployed for testing in this paper is constructed to mimic the $N=47$ pulsars from the 12.5-year NANOGrav data release, whose sky positions are plotted in Figure \ref{fig:pulsar_distrib}. The construction recipe is described in full in Section \ref{sec:synt_pta}. In this appendix, we record a complete list of the pulsar parameters in Table \ref{tab:pulsar_values}, to assist the interested reader in implementing the analysis scheme and reproducing the test results in Sections \ref{sec:rep_example}--\ref{sec:bias_and_identifiability}.

\begin{table*}
\begin{tabular}{lccccc}
	\toprule
	JNAME &    $f_{\rm em} (t_1)$ (Hz) &  $\dot{f}^{(n)}_{\rm em}(t_1)\times 10^{16}$ (s$^{-2}$) &  $\alpha$ (deg) &  $\delta$ (deg) &  $d$ (kpc) \\
	\midrule
	J0023+0923 & 327.8 &      -12.3 &   5.8 &    9.4 &   1.8 \\
	J0030+0451 & 205.5 &       -4.3 &   7.6 &    4.9 &   0.3 \\
	J0340+4130 & 303.1 &       -6.5 &  55.1 &   41.5 &   1.6 \\
	J0613-0200 & 326.6 &      -10.2 &  93.4 &   -2.0 &   0.9 \\
	J0636+5128 & 348.6 &       -4.2 &  99.0 &   51.5 &   0.7 \\
	J0645+5158 & 112.9 &       -0.6 & 101.5 &   52.0 &   1.2 \\
	J0740+6620 & 346.5 &      -14.6 & 115.2 &   66.3 &   1.1 \\
	J0931-1902 & 215.6 &       -1.7 & 142.8 &  -19.0 &   3.7 \\
	J1012+5307 & 190.3 &       -6.2 & 153.1 &   53.1 &   0.7 \\
	J1024-0719 & 193.7 &       -7.0 & 156.2 &   -7.3 &   1.2 \\
	J1125+7819 & 238.0 &       -3.9 & 171.5 &   78.3 &   0.9 \\
	J1453+1902 & 172.6 &       -3.5 & 223.4 &   19.0 &   1.3 \\
	J1455-3330 & 125.2 &       -3.8 & 223.9 &  -33.5 &   0.7 \\
	J1600-3053 & 277.9 &       -7.3 & 240.2 &  -30.9 &   1.9 \\
	J1614-2230 & 317.4 &       -9.7 & 243.7 &  -22.5 &   0.7 \\
	J1640+2224 & 316.1 &       -2.8 & 250.1 &   22.4 &   1.5 \\
	J1643-1224 & 216.4 &       -8.6 & 250.9 &  -12.4 &   0.7 \\
	J1713+0747 & 218.8 &       -4.1 & 258.5 &    7.8 &   1.3 \\
	J1738+0333 & 170.9 &       -7.0 & 264.7 &    3.6 &   1.5 \\
	J1741+1351 & 266.9 &      -21.5 & 265.4 &   13.9 &   1.7 \\
	J1744-1134 & 245.4 &       -5.4 & 266.1 &  -11.6 &   0.4 \\
	J1747-4036 & 607.7 &      -48.5 & 267.0 &  -40.6 &   7.1 \\
	J1832-0836 & 367.8 &      -11.2 & 278.1 &   -8.6 &   2.1 \\
	J1853+1303 & 244.4 &       -5.2 & 283.5 &   13.1 &   2.1 \\
	J1857+0943 & 186.5 &       -6.2 & 284.4 &    9.7 &   1.2 \\
	J1903+0327 & 465.1 &      -40.7 & 285.8 &    3.5 &   7.0 \\
	J1909-3744 & 339.3 &      -16.1 & 287.4 &  -37.7 &   1.1 \\
	J1910+1256 & 200.7 &       -3.9 & 287.5 &   12.9 &   1.5 \\
	J1911+1347 & 216.2 &       -7.9 & 288.0 &   13.8 &   1.4 \\
	J1918-0642 & 130.8 &       -4.4 & 289.7 &   -6.7 &   1.1 \\
	J1923+2515 & 264.0 &       -6.7 & 290.8 &   25.3 &   1.2 \\
	J1939+2134 & 641.9 &     -433.1 & 294.9 &   21.6 &   3.5 \\
	J1944+0907 & 192.9 &       -6.4 & 296.0 &    9.1 &   1.2 \\
	J1946+3417 & 315.4 &       -3.1 & 296.6 &   34.3 &   6.9 \\
	J1955+2908 & 163.0 &       -7.9 & 298.9 &   29.1 &   6.3 \\
	J2010-1323 & 191.5 &       -1.8 & 302.7 &  -13.4 &   2.4 \\
	J2017+0603 & 345.3 &       -9.5 & 304.3 &    6.1 &   1.4 \\
	J2033+1734 & 168.1 &       -3.1 & 308.4 &   17.6 &   1.7 \\
	J2043+1711 & 420.2 &       -9.3 & 310.8 &   17.2 &   1.4 \\
	J2145-0750 &  62.3 &       -1.2 & 326.5 &   -7.8 &   0.7 \\
	J2214+3000 & 320.6 &      -15.1 & 333.7 &   30.0 &   0.6 \\
	J2229+2643 & 335.8 &       -1.7 & 337.5 &   26.7 &   1.8 \\
	J2234+0611 & 279.6 &       -9.4 & 338.6 &    6.2 &   1.0 \\
	J2234+0944 & 275.7 &      -15.3 & 338.7 &    9.7 &   1.6 \\
	J2302+4442 & 192.6 &       -5.1 & 345.7 &   44.7 &   0.9 \\
	J2317+1439 & 290.3 &       -2.0 & 349.3 &   14.7 &   1.7 \\
	J2322+2057 & 208.0 &       -4.2 & 350.6 &   21.0 &   1.0 \\
	\bottomrule
\end{tabular}
\caption{List of the fiducial pulsar parameters for the 47 pulsars used to construct the synthetic PTA in Section \ref{sec:synt_pta}. The right ascension and declination of an individual pulsar in J2000 coordinates are labelled by $\alpha$ and $\delta$ respectively. Parameter values are obtained from the ATNF pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}.}
\label{tab:pulsar_values}
\end{table*}

\section{Dispersion of ${\theta}_{\rm gw}$ estimates}\label{sec:wasserstein}
Every random realization of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ leads to a different ${\boldsymbol{\theta}}_{\rm gw}$ posterior, when the synthetic data are analysed according to the procedure in Section \ref{sec:methodsummary}. This cosmic variance is unavoidable. It is a key source of uncertainty, when the analysis is performed on real astronomical data, where the specific random realizations of $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$ are unknown. To quantify the cosmic variance, one needs a metric to describe the distance between two posteriors (i.e.\ how similar they are). Many valid metrics of this sort exist, including those related to the Kolmogorov-Smirnoff test \citep{corder2014nonparametric}. In this paper, we use the WD \citep{Wasserstein,Villani2009} which is popular in machine learning \citep[e.g.][]{2017arXiv170107875A} and other domains. In Section \ref{sec:appendix_overview_WD} we define the WD and summarize its main properties. In Section \ref{sec:appendix_WD_results} we present for reproducibility the WD calculated between every pair of posteriors for each static parameter in $\boldsymbol{\theta}_{\rm gw}$ across the $10^3$ noise realisations of Section  \ref{sec:multiple_noise}. The results are summarised via the median values reported in Table \ref{tab:Wasserstein}.
